---
title: "IS4300 Final Project"
author: "Cabot Boyd"
date: '2022-04-27'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

  We will be examining a data set found on Github called "Yearly batting records 
for all major league baseball players." It contains a total of 21,699 records, covering 1,228 players from 1871 to 2007 and only contains players with more 
than 15 seasons played. 

The data contained in this .csv is various batting related statistics, such as 
at bats, home runs, stolen bases, and hits by pitch. 

This report will be examining Gary Sheffield's case for making the MLB Hall of
Fame. We will doing this by comparing his offensive performance to other Hall of
Famers from the same era.

```{r,echo=FALSE}
library (tidyverse)
library (dslabs)
library (readr)
urlfile="https://vincentarelbundock.github.io/Rdatasets/csv/plyr/baseball.csv"
baseball<-read_csv(url(urlfile))
str(baseball)
```
## Visualizations

The first visual is a representation of Gary Sheffield's batting average over
his career compared to HOFer Ken Griffey Jr., who played over the same span.

Figure 1
```{r,echo=FALSE}
# We need to compare a few offensive stats
# The first is batting average. Because there is no batting average variable in
# this data set, we are going to have to create one using the mutate function

avg <- baseball %>% filter(id == "sheffga01" |  id == "griffke02") %>% 
  mutate(ba = (h)/ab) %>% select(id,year,h,ab,ba)
# This creates a data frame that filters out the two players we want to examine
# and create a new column that creates the batting average statistic
batavg <- avg %>% ggplot() + geom_line(aes(year,ba,color=id),size=1.2) + 
  labs(title = "Career Batting Averages", 
       subtitle = "Ken Griffey Jr. v. Gary Sheffield", x = "Year", 
       y = "Batting Average") + 
  scale_colour_discrete(name = "Players",labels = c("Griffey Jr.","Sheffield")) +
  geom_hline(yintercept = 0.25) + 
  geom_text(aes(1999,0.245),label = "League Average",size = 4)
batavg
```

Next, we will look at another HOFer and teammate on the San Diego Padres, Tony
Gwynn, and compare their home run numbers.

Figure 2
```{r,echo=FALSE}
# Now we will compare Sheffield's home run numbers to another HOFer and teammate,
# Tony Gwynn. 20+ home runs in a season would be considered elite.

hits <- baseball %>% filter(id == "sheffga01" | id == "gwynnto01") %>% 
  select(id,year,h,X2b,X3b,hr)
# This creates a data frame that filters out just the two players we want to 
# examine
homeruns <- ggplot(hits,aes(fill=id, y=hr, x=year)) + 
  geom_bar(position="identity", stat = "identity", alpha = 0.6) + 
  labs(title = "Career Home Runs", subtitle = "Gary Sheffield v. Tony Gwynn", 
       x = "Year", y = "Homeruns") + geom_hline(yintercept = 20) +
  geom_text(aes(1985,22),label = "Elite Level",size = 4)
homeruns
```

Finally we will compare Sheffield's on base plus slugging percent, or OPS,
compared to another Hall of Fame Inductee who played in that era, Rickey
Henderson.

Figure 3
```{r,echo=FALSE}
# Finally we will look at OPS, or On base Plus Slugging, which represents the 
# offensive production of a player. An elite level for OPS is 0.450+. We will
# compare Sheffield to HOFer Rickey Henderson.

plus <- baseball %>% mutate(ops=((h+X2b+X3b+hr+bb+hbp)/(ab+hbp+sf))) %>% 
  filter(id == "sheffga01" | id == "henderi01")
# We have to establish a data frame that adds a column for OPS and filter for
# the two players we are looking at
onbaseps <- plus %>% ggplot() + geom_line(aes(year,ops,color=id),size=1.2) +
  labs(title = "Career Onbase Plus Slugging", 
       subtitle = "Gary Sheffield v. Rickey Henderson", x = "Year", y = "OPS") +
  scale_colour_discrete(name = "Players",labels = c("Henderson","Sheffield")) +
  geom_hline(yintercept = 0.450) + 
  geom_text(aes(1985,0.465),label = "Elite Level",size = 4)
onbaseps
```

## Analysis

Looking at Sheffield's numbers compared to Hall of Fame Inductees, we can see
that he has comparable numbers. Sheffield played at or above an elite level on
the offensive side of the ball throughout his entire career.

```{r,echo=FALSE}
batavg
```

Looking at Figure 1, we can see that Sheffield's batting average was comparable
to Ken Griffey Jr., and well above the league average of 0.250 for several 
seasons. Sheffield had a career high batting average of .330 and a career average
of .248, with an injured 1991 season bringing the average down.

```{r,echo=FALSE}
homeruns
```

Figure 2 shows that Sheffield had 12 seasons with more than 20 home runs in the
season. He also had more home runs than Tony Gwynn, a Hall of Famer. Sheffield
hit 509 homeruns over his career, being 1 of 28 players to hit 500 home runs. 14
members of the 500 HR club are in the Hall of Fame. Of the other half, 2 players
are active, and 11 players have been accused or found guilty of using steroids.
That leaves only Gary Sheffield.

```{r,echo=FALSE}
onbaseps
```

Examining Figure 3 tells us a lot. Rickey Henderson did not play all the same
seasons but they had over a decade of overlap. While Henderson, a Hall of Famer,
had a more consistent career in terms of OPS, Sheffield still had many seasons
with OPS at an elite level, with only 3 seasons below .450. Sheffield also ranks
63rd in career OPS in the MLB, just below Ken Griffey Jr. 

## Conclusion

Gary Sheffield has appeared on the Hall of Fame ballot for eight years, with only
two more years remaining before he gets left off the ballot and misses out on
joining the greats in Cooperstown. Using only the above statistics and visuals,
you can see that Sheffield makes a compelling case for entering the halls of
Cooperstown. He has comparable numbers to Hall of Famers and has played at an
elite level on offense for the majority of his career. If I had a vote, Gary
Sheffield, the 9 time All-Star, 5 time Silver Slugger, and World Champ, would 
definitely make it into the Hall of Fame.